Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2023-11-14 Delaneyetal2023b.tex

\title{Natural variations in water discharge quantity have an inconsistent impact on sediment transport capacity in subglacial channels\mauro{\\Title does not work.  I don't think ``inconsistent'' can be used like this. Maybe:}\\
Subglacial and subaerial fluvial sediment transport respond differently to discharge variations
}

@@

@@

\begin{abstract}  Sediment transport capacity in subglacial and subaerial channels depends on  water velocity and channel width, which together control the total shear stress exerted on the channel bottom.
  In subaerial channels water discharge variations are accommodated largely by changes in flow depth and width.
  While in subglacial channels, water discharge changes primarily affect water velocity due to slow-evolving channel geometry.
  Parameterizations of these subglacial and subaerial hydraulics are applied to hydrographs from an Alpine glacier and the Greenland Ice Sheet revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels.
  Results show that high subglacial sediment transport rates can occur across a wide range of water discharges.
%  Slowly evolving channel width accommodates some variability in subglacial channels. \mauro{not sure what that sentence should do}
  Despite differences in response to water discharge, sediment discharge measurements generally correlate better with subaerial hydraulic parameters.
  This implies that other factors like sediment access or proglacial processes might be at play that affect sediment transport capacity differently in subglacial and subaerial channels.
\end{abstract}

@@

In a transport-limited regime, sediment discharge tracks sediment transport capacity, defined as the amount of sediment the water can carry.
Sediment transport capacity depends on the shear stress between water and the sediment it flows  over \cite{shields1936,meyer1948,engelund1967} along with the width of the channel bottom $w$ over which to mobilize sediment.
The shear stress $\tau$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau}
    \tau_t \propto v^2.
  \end{equation}
\end{linenomath*}
\mauro{I removed the width from here as I found it odd to introduce that here.}
%
Following mass conservation, the velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area. \mauro{I find it odd to have formulas in the Intro}

@@

Across the range of channel shape parameters and friction factors  examined, variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one (Figure~\ref{fig:multi_run}).
In \alpine{}, the variability in width-integrated shear stress is comparable between the two systems, due to the variable water discharge.
However, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains higher in the subglacial system across the time aggregation periods.

@@

\mauro{You often use ``subglacial hydraulics''.  I'm not sure what that means. For instance in the next sentence.  Isn't discharge part of hydraulics too?  Maybe you could define a bunch of equations as the subglacial hydraulic model and subaerial hydraulic model and then talk of that?}
%
The stronger correlation between subglacial hydraulics and water discharge in \icesheet{} (Figure~\ref{fig:Qw_vari}a-f) could result from the channel being closer to equilibrium between variations in the channels' wetter area and water discharge with smaller diurnal variations in amplitude (Figures~\ref{fig:model_outs} and ~\ref{fig:Qw_vari}) \mauro{needs rewriting}.
Conversely, the correlation between hydraulics and water discharge in \alpine is smaller due to the relatively larger discharge variations on short time scales.
In both scenarios, hydraulics could \mauro{``could''?  Does it or does it not?} trend toward equilibrium by averaging the discharge over $1-5$ days, when the variability in subglacial channels approaches that of subaerial ones (Figure~\ref{fig:multi_run}).

In subglacial channels, peaks in water velocity, shear stress and width-integrated shear stress occur with  fast rates of water discharge increase relative to the channel's growth rate (Figure~\ref{fig:model_outs}).
In turn, subglacial channels grow with increasing water discharge, effectively reducing sediment transport capacity with the stabilization of water discharge at its peak.
Similar feedback occurs in subaerial systems \cite{phillips2016}, albeit at timescales longer than days.  \mauro{\citeA{gimbert2016}, which you cited here, says nothing of subaerial flow, or do they?}

\subsection{Records of sediment discharge  from glaciers}
\mauro{This section header sounds like this will be about measurements only.}

@@

First, these measurement stations lie $0.5$ to $2$\,\unit{km} downstream of the glacier terminus \cite{cowton2012,felix2022}, a long enough distance for the pressurized subglacial regime to transition to the open channel subaerial one.
The transition of subglacial to subaerial conditions typically occur still underneath glacier but close to the terminus \cite{perolo2018}, where closure rates of the subglacial channels are slow and open channel flow condition prevail \cite{rothlisberger1972}.
Furthermore, if a transport-limited regime persists at the glacier margin, then sediment can be easily mobilized, and subaerial processes could be represented just a short distance downstream \mauro{Not sure what that sentence means}.

\mauro{I perceived this paragraph pretty out of the blue.}
The consistent relationship between sediment discharge capacity and open channel flow conditions \mauro{What? How do you compare those two things?} could be pronounced $\sim20$\,\unit{km} downstream of the Leverett site, where a strong correlation persists between sediment plume size and the river's water discharge into the Kangerlussuaq fjord \cite{mcgrath2010}.
In contrast, in marine-terminating glacier catchments a less consistent relationship may occur between water discharge or melt extent and sediment plume size \cite{tedstone2012}.
Here, ocean water pressurizes subglacial channels at the ice front \cite<e.g.>{how2017}, so the minimal correlation could result from the inconsistent relationship between subglacial sediment transport capacity and water discharge (Figure~\ref{fig:Qw_vari}).

Second, the sediment discharge records could represent processes such as increased sediment accessibility when melt extends up the glacier, thereby increasing sediment and water discharge \cite<e.g.>{vergara2022}.
A supply-limited regime could well persist at many glaciers, because the sediment's threshold of motion can be reached more frequently and across a range of water discharges, compared to subaerial systems (Figure~\ref{fig:Qw_vari}).
Sediment exhaustion occurring through increased variability may also explain the stronger dependence of sediment discharge from the Greenland Ice Sheet on basal shear stress, a proxy for bedrock erosion, rather than glacier melt \cite{overeem2017}.
Note, however, the absence of lags between the observed sediment and water discharge \mauro{reference a figure here?} may suggest that sediment is mobilized close to the measurement stations, as opposed to distal areas of the glacier bed as would be expected in a supply-limited regime \cite<Figure~\ref{fig:model_outs}\, a and e; >{williams1989}.

Regardless of the increased performance of water discharge compared to subglacial sediment transport capacity in the observations \mauro{does not parse}, outputs suggest that additional processes and erosional hiatuses may further complicate signals of sediment discharge from glaciers in response to climate \cite{jansson2005,ganti2016}.
Sediment discharge records have been used to establish the relationship, or lack thereof, between climate forcing and glacial erosion \cite<e.g.>{koppes2009a,ganti2016,willenbring2016,mariotti2021}.
The variable relationship between water discharge and subglacial sediment transport capacity presented here mostly applies to the daily- to weekly-time scales controlling the size of subglacial channels (Figure~\ref{fig:multi_run}).
Yet, identifying climatic signals in sediment transport from the transport-limited component of glacier-sedimentary systems may require higher thresholds of water discharge compared to subaerial systems when water discharge varies substantially over short timescales \cite<Figure~\ref{fig:Qw_vari}; >{tofelde2021}. \mauro{I don't understand the last two sentences.}

@@

@@

The sediment transport capacity of a channel is driven by its width and the shear stress exerted by the flowing water, which, in turn, is proportional to the square of the flow velocity.
Subaerial channels alter both their channel width and water velocity immediately in response to changing water discharge.
In contrast, pressurized subglacial channels largely accommodate rapidly changing water discharge by altering water velocity but not their size; the latter, in contrast, only responds over the course of several days to changes in discharge.
Thus the response of sediment transport capacity is much more sensitive to changes in water discharge in subglacial channels compared to subaerial ones.
Despite these differences, parameterizations of subaerial hydraulics in two glacierized catchments generally represent observed sediment discharge variations better than subglacial hydraulics \mauro{This sentence does not quite parse.  Why ``despite''?  Also state that we found this.}

Implications of the presented findings include:
\begin{enumerate}
\item the inconsistent \mauro{``inconsistent'' does not work for me here. What do you mean? More non-linear?} relationship between water discharge and sediment transport capacity underneath glaciers creates larger variability in sediment transport capacity compared to river systems;
\item the timing of peak sediment transport capacity below glaciers generally coincides with large rates of increase in water discharge, as opposed to the maximum water discharge;
\item the timing of peak sediment transport capacity in subaerial channels coincides with the maximum water discharge;
\item a consistent \mauro{and what does ``consistent'' mean?} relationship between water and sediment discharge may represent subglacial sediment access or open channel hydraulics near the glacier margin, instead of subglacial sediment transport capacity.
\end{enumerate}

@@

Here three kinds of drainage regimes are examined, namely, subaerial channels, steady-state R-channels \cite{rothlisberger1972}, and pipe-flow (i.e. R-channels which do not have time to adjust their size to discharge conditions).
Here we show how sediment transport capacity scales with respect to a given water discharge and hydraulic gradient for three different transport formulas: MPM \cite{meyer1948}, EH \cite{engelund1967}, and Bagnold \cite{bagnold1980}; additionally width-integrated shear stress is assessed, as used as proxy for sediment transport throughout the manuscript.

>>>>

\title{Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations}

@@

change subglacial hydraulics to subglacial model outputs??

@@

\begin{abstract}  Sediment transport capacity in subglacial and subaerial channels depends on  water velocity and channel width, which together control the total shear stress exerted on the channel bottom.
  In subaerial channels water discharge variations are accommodated largely by changes in flow depth and width.
  While in subglacial channels, water discharge changes primarily affect water velocity due to slow-evolving channel geometry.
  Parameterizations of these subglacial and subaerial hydraulics are applied to hydrographs from an Alpine glacier and the Greenland Ice Sheet revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels.
  Results show that high subglacial sediment transport rates can occur across a wide range of water discharges.
  Slowly evolving channel width accommodates some variability in subglacial channels. \mauro{not sure what that sentence should do}
  Despite differences in response to water discharge, sediment discharge measurements generally correlate better with subaerial hydraulic parameters.
  This implies that other factors like sediment access or proglacial processes might be at play that affect sediment transport capacity differently in subglacial and subaerial channels.
\end{abstract}

@@

In a transport-limited regime, sediment discharge tracks sediment transport capacity, defined as the amount of sediment the water can carry.
Sediment transport capacity depends on the shear stress between water and the sediment it flows  over \cite{shields1936,meyer1948,engelund1967} along with the width of the channel bottom $w$ over which to mobilize sediment.
The shear stress $\tau$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau}
    \tau_t \propto w\,v^2.
  \end{equation}
\end{linenomath*}
\mauro{I removed the width from here as I found it odd to introduce that here.}
%
Following mass conservation, the velocity of the water flowing through a  channel is
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area. \mauro{I find it odd to have formulas in the Intro}

@@

Across the range of channel shape parameters and friction factors  examined, variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one.
In \alpine{}, the variability in width-integrated shear stress is comparable between the two systems, due to the variable water discharge (Figure~\ref{fig:multi_run}).

However, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains higher in the subglacial system across the time aggregation periods.

@@

\mauro{You often use ``subglacial hydraulics''.  I'm not sure what that means. For instance in the next sentence.  Isn't discharge part of hydraulics too?  Maybe you could define a bunch of equations as the subglacial hydraulic model and subaerial hydraulic model and then talk of that?}
%
The stronger correlation between subglacial hydraulics and water discharge in \icesheet{} (Figure~\ref{fig:Qw_vari}\,a-f) could result from the channel being closer to equilibrium between variations in the channels' wetter area and water discharge with smaller diurnal variations in amplitude (Figures~\ref{fig:model_outs} and ~\ref{fig:Qw_vari}) \mauro{needs rewriting}.
Conversely, the correlation between hydraulics and water discharge in \alpine is smaller due to the relatively larger discharge variations on short time scales.
In both scenarios, the variability in subglacial channels approaches that of subaerial ones  over $1-5$ days (Figure~\ref{fig:multi_run}), suggesting a parity between the two channel types might be approached over this period.

In subglacial channels, peaks in water velocity, shear stress and width-integrated shear stress occur with  fast rates of water discharge increase relative to the channel's growth rate (Figure~\ref{fig:model_outs}).
In turn, subglacial channels grow with increasing water discharge, effectively reducing sediment transport capacity with the stabilization of water discharge at its peak.
Similar feedback occurs in subaerial systems \cite{phillips2016}, albeit at timescales longer than days in the subglacial system.

\subsection{Effect of variable sediment transport capacity on sediment discharge  from glaciers}
\mauro{This section header sounds like this will be about measurements only.}

@@

First, these measurement stations lie $0.5$ to $2$\,\unit{km} downstream of the glacier terminus \cite{cowton2012,felix2022}, a long enough distance for the pressurized subglacial regime to transition to the open channel subaerial one.
The transition of subglacial to subaerial conditions typically occur still underneath glacier but close to the terminus \cite{perolo2018}, where closure rates of the subglacial channels are slow and open channel flow condition prevail \cite{rothlisberger1972}.
Furthermore, if abundant sediment persists in the proglacial area, then it can be easily mobilized, and subaerial processes could be represented just a short distance downstream \mauro{Not sure what that sentence means}.

\mauro{I perceived this paragraph pretty out of the blue.}
The consistent relationship between sediment discharge capacity in  open channel flow conditions could be pronounced $\sim20$\,\unit{km} downstream of the Leverett site, where a strong correlation persists between sediment plume size and the river's water discharge into the Kangerlussuaq fjord \cite{mcgrath2010}.
In contrast, in marine-terminating glacier catchments a less consistent relationship may occur between water discharge or melt extent and sediment plume size \cite{tedstone2012}.
Here, ocean water pressurizes subglacial channels at the ice front \cite<e.g.>{how2017}, so the minimal correlation could result from the inconsistent relationship between subglacial sediment transport capacity and water discharge (Figure~\ref{fig:Qw_vari}).

Second, the sediment discharge records could represent processes such as increased sediment accessibility when melt extends up the glacier, thereby increasing sediment and water discharge \cite<e.g.>{vergara2022}.
A supply-limited regime could well persist at many glaciers, because the sediment's threshold of motion can be reached more frequently and across a range of water discharges, compared to subaerial systems (Figure~\ref{fig:Qw_vari}).
Sediment exhaustion occurring through increased variability may also explain the stronger dependence of sediment discharge from the Greenland Ice Sheet on basal shear stress, a proxy for bedrock erosion, rather than glacier melt \cite{overeem2017}.
Note, however, the absence of lags between the observed sediment and water discharge (Figure~\ref{fig:model_outs}\, a and e) may suggest that sediment is mobilized close to the measurement stations, as opposed to distal areas of the glacier bed as would be expected in a supply-limited regime \cite{williams1989}.

Regardless of the increased performance of water discharge compared to subglacial sediment transport capacity in representing the observations,  suggest that additional processes and erosional hiatuses may further complicate signals of sediment discharge from glaciers in response to climate \cite{jansson2005,ganti2016}.
Sediment discharge records have been used to establish the relationship, or lack thereof, between climate forcing and glacial erosion \cite<e.g.>{koppes2009a,ganti2016,willenbring2016,mariotti2021}.
The variable relationship between water discharge and subglacial sediment transport capacity presented here mostly applies to the daily- to weekly-time scales controlling the size of subglacial channels (Figure~\ref{fig:multi_run}).
Yet, identifying climatic effects in sediment transport from the transport-limited component of glacier-sedimentary systems may require higher thresholds of water discharge compared to subaerial systems when water discharge varies substantially over short timescales \cite<Figure~\ref{fig:Qw_vari}; >{tofelde2021}. \mauro{I don't understand the last two sentences.}

@@

\ian{move this shit up or delete}
\subsection{Comparison of Alpine and Arctic hydrological forcing}

@@

The sediment transport capacity of a channel is driven by its width and the shear stress exerted by the flowing water, which, in turn, is proportional to the square of the flow velocity.
Subaerial channels alter both their channel width and water velocity immediately in response to changing water discharge.
In contrast, pressurized subglacial channels largely accommodate rapidly changing water discharge by altering water velocity.
Their size, in contrast, responds over the course of several days to changes in discharge.
Thus the response of sediment transport capacity is much more sensitive to changes in water discharge in subglacial channels compared to subaerial ones.
%Despite these differences, parameterizations of subaerial hydraulics in two glacierized catchments generally represent observed sediment discharge variations better than subglacial hydraulics \mauro{This sentence does not quite parse.  Why ``despite''?  Also state that we found this.}

Implications of the presented findings include:
\begin{enumerate}
\item the changing relationship between water discharge and sediment transport capacity underneath glaciers creates larger variability in sediment transport capacity compared to river systems;
\item the timing of peak sediment transport capacity below glaciers generally coincides with large rates of increase in water discharge, as opposed to the maximum water discharge, as would occur in pipe flow of a fixed size;
\item the timing of peak sediment transport capacity in subaerial channels coincides with the maximum water discharge;
\item a steady relationship between water and sediment discharge may represent subglacial sediment access or open channel hydraulics near the glacier margin, instead of subglacial sediment transport capacity.
\end{enumerate}

@@

Here three kinds of drainage regimes are examined, namely, subaerial channels, steady-state R-channels \cite{rothlisberger1972}, and pipe-flow (i.e. R-channels which do not have time to adjust their size to discharge conditions).
Here we show how sediment transport capacity scales with respect to a given water discharge and hydraulic gradient for three different transport formulas: Meyer-Peter M\"uller  \cite<MPM;>{meyer1948}, Engelund and Hansen \cite<EH;>{engelund1967}, and Bagnold \cite{bagnold1980}; additionally width-integrated shear stress is assessed, as used as proxy for sediment transport throughout the manuscript.
